---
title: "p8105_hw2_bac2214"
author: "Brianna Carnagie"
date: "2023-10-02"
output: github_document
---

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(readxl)
library(dplyr)
```

# Problem 1
#### Cleaning the first dataset
```{r, message=FALSE}
pols_month_df = 
  read_csv("data/pols-month.csv") |> 
  janitor::clean_names() |> 
  separate_wider_delim(mon, "-", names = c("year","month","day")) |> 
  mutate(
    month = case_match(
    month,
    "01" ~ "jan",
    "02" ~ "feb",
    "03" ~ "mar",
    "04" ~ "apr",
    "05" ~ "may",
    "06" ~ "jun",
    "07" ~ "jul",
    "08" ~ "aug",
    "09" ~ "sep",
    "10" ~ "oct",
    "11" ~ "nov",
    "12" ~ "dec"), year = as.double(year)) |> 
  mutate(president = case_when(
     prez_dem == 1 ~ "dem",
    prez_gop == 1 ~ "gop",
    prez_gop == 2 ~ "gop"
  )) |> select(-prez_dem, -prez_gop, -day) 
``` 

#### Cleaning the second dataset
```{r, message=FALSE}
snp_df = 
  read_csv("data/snp.csv") |> 
  janitor::clean_names() |> 
  separate_wider_delim(date, "/", names = c("month", "day", "year")) |> 
  select(year, month,-day, close,) |> 
  mutate(
    month = case_match(
    month,
    "1" ~ "jan",
    "2" ~ "feb",
    "3" ~ "mar",
    "4" ~ "apr",
    "5" ~ "may",
    "6" ~ "jun",
    "7" ~ "jul",
    "8" ~ "aug",
    "9" ~ "sep",
    "10" ~ "oct",
    "11" ~ "nov",
    "12" ~ "dec")) |> mutate(year = as.double(year), year = year + 2000)
```

#### Cleaning the last dataset 
```{r, message=FALSE}
unemployment_df = 
  read_csv("data/unemployment.csv") |> 
  janitor::clean_names() |> 
  pivot_longer(jan:dec,
               names_to = "month",
               values_to = "unemployment_percentage") 
```
  
#### Merging the dataset
```{r}
fivethirtyeight_merge = 
  left_join(pols_month_df, snp_df) |>
  left_join(x = _, y = unemployment_df)
```

#### Making variables about the dataset
```{r}
min_year <- fivethirtyeight_merge |> 
  select(year) |> 
  pull() |> 
  min()

max_year <- fivethirtyeight_merge |> 
  select(year) |> 
  pull() |> 
  max()

num_rows = nrow(fivethirtyeight_merge)
num_col = ncol(fivethirtyeight_merge)
```

The data contains `r num_col` columns and `r num_rows` rows . I see that the first year of data is `r min_year`  while the most recent year is `r max_year`. The `month` and `year` variables were very important for merging the 3 datasets together. An abbreviation wasn't really necessary in this case, but it was good to keep consistent naming conventions across all 3 datasets.

## Problem 2

#### Cleaning the first dataset
```{r}
mr_trashwheel_df =
  read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N586") |> janitor::clean_names() |> mutate(homes_powered = (weight_tons * 500)/30, trashwheel_name = "mr_trashwheel", year = as.double(year)) |> relocate(trashwheel_name)
```

#### Cleaning the second dataset
```{r}
professor_trashwheel_df = 
  read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M108") |> janitor::clean_names() |> mutate(homes_powered = (weight_tons * 500)/30, trashwheel_name = "professor_trashwheel")|> relocate(trashwheel_name)
```

#### Cleaning the last dataset
```{r}
gwynnda_trashwheel_df = 
  read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", range = "A2:L157") |>     janitor::clean_names() |> mutate(homes_powered = (weight_tons * 500)/30, trashwheel_name = "gwynnda_trashwheel") |> relocate(trashwheel_name)
```

#### Binding all 3 datasets together
```{r}
trashwheel_bind_df = 
  bind_rows(mr_trashwheel_df, professor_trashwheel_df, gwynnda_trashwheel_df)
```

 
#### Making variables about the dataset
```{r}
num_rows2 = nrow(trashwheel_bind_df)
num_col2 = ncol(trashwheel_bind_df)

na_count = sum(is.na(trashwheel_bind_df$plastic_bags))

cigbutts_count = trashwheel_bind_df |> 
  select(plastic_bags) |> 
  pull() |> 
  sum(na.rm = TRUE)
```

There are `r num_col2` columns and `r num_rows2` observations. I noticed that there are `r na_count` in the `plastic bag` variable due to the fact that `mr_trashwheel_df` and `professor_trashwheel_df` did not collect data for this variable. Cigarette butts seem be a huge contributor of trash, with all 3 trashwheels collecting a total of `r cigbutts_count` butts total. 


> For available data, what was the total weight of trash collected by Professor Trash Wheel?
```{r}
professortrashwheel_total_wt = trashwheel_bind_df |> 
  select(trashwheel_name, weight_tons) |> filter(trashwheel_name == "professor_trashwheel") |> 
  pull("weight_tons") |> 
  sum(na.rm = TRUE)
```

The total weight of trash collected by Professor Trash Wheel is `r professortrashwheel_total_wt` tons.


> What was the total number of cigarette butts collected by Gwynnda in July of 2021?
```{r}
gwynndatrashwheel_total_cigbutts = subset(trashwheel_bind_df,
  trashwheel_name == "gwynnda_trashwheel" &
  month == "July" &
  year == 2021) |> 
select(cigarette_butts) |> 
sum(na.rm = TRUE)  
```

The total number of cigarette butts collected by Gwynnda in July of 2021 was `r gwynndatrashwheel_total_cigbutts`.


